---
title: "Homework 10"
author: "Blue Whales"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("network")
#install.packages("igraph")
library(readr)
library(dplyr)
library(tidyverse)
library(network)
library(igraph)
```

**Group Members:** Arvind Bala, Ben Brikman, Jimena Velazquez, Minglu Xu, Eva Zhang

# Question 1
We prepared the data for the network analysis by creating a separate edge and node list.

```{r, echo = F, include = F}
# Load data, create nodelist and edgelist

hw10_data <- read_csv("hw10_data.csv")
hw10_data = hw10_data[-1,]
hw10_data <- hw10_data[!is.na(hw10_data$Q2), ]

nodelist <- select(hw10_data,Q2,Q4) %>% 
  rename('ID' = Q2, 'Extraversion' = Q4) %>% 
  transform(ID = as.numeric(ID)) %>% 
  add_row(ID = 3, Extraversion = NA) %>% 
  add_row(ID = 4, Extraversion = NA) %>%
  add_row(ID = 8, Extraversion = NA) %>%
  add_row(ID = 9, Extraversion = NA) %>%
  add_row(ID = 10, Extraversion = NA) %>%
  add_row(ID = 11, Extraversion = NA) %>%
  add_row(ID = 12, Extraversion = NA) %>%
  add_row(ID = 13, Extraversion = NA) %>%
  add_row(ID = 14, Extraversion = NA) %>%
  add_row(ID = 16, Extraversion = NA) %>%
  add_row(ID = 19, Extraversion = NA) %>%
  add_row(ID = 21, Extraversion = NA) %>%
  add_row(ID = 22, Extraversion = NA) %>%
  add_row(ID = 25, Extraversion = NA) %>%
  add_row(ID = 26, Extraversion = NA) %>%
  add_row(ID = 27, Extraversion = NA) %>%
  add_row(ID = 29, Extraversion = NA) %>%
  add_row(ID = 37, Extraversion = NA) %>%
  add_row(ID = 39, Extraversion = NA) 

edgelist <- select(hw10_data,-Q4)
edgelist <- edgelist %>% 
  pivot_longer(!Q2, names_to = 'in-tie', values_to = 'tie-num', values_drop_na = TRUE) %>% 
  rename('out-tie' = Q2) %>% 
  separate('tie-num', into = c("1", "2","3","4","5"), sep = ",") %>% 
  pivot_longer('1':'5', names_to = 'number', values_to = 'tie-type', values_drop_na = TRUE) %>% 
  select('out-tie', 'in-tie', 'tie-type') 
edgelist$`in-tie` <- gsub("^.{0,3}", "", edgelist$`in-tie`)
```

# Question 2
*Calculate measures of centrality and similarity in networks depending on tie-type?
Which tie type generates the most similar network in terms of introversion? 
Which tie type is the most dissimilar?*

**Network Visualization by Tie Type**

We start our analysis by visualizing the overall network and the 5 networks
filtered by the 5 tie types. It's interesting to observe that
networks based on tie type 3, 4, 5 have relatively more isolated nodes. These ties 
describes friendships and relations of asking/being asked for help. Perhaps 
remote learning has impacted people's ability to make friends in the program.

```{r, echo = F}
detach(package:network)
newnetwork <- graph_from_data_frame(d = edgelist, vertices = nodelist, directed = TRUE)
# Plot overall network of all ties
plot(newnetwork, edge.arrow.size = 0.1, vertex.size = 3, main = "Overall Network")

# Plot overall network varying node size based on number of ties
#deg <- degree(newnetwork, mode="all")
#plot(newnetwork, vertex.size=deg/10, edge.arrow.size = 0.1, main = "Overall Network")

# filter by all 5 tie types
edgelist_1 <- edgelist %>% filter(`tie-type` == 1)
edgelist_2 <- edgelist %>% filter(`tie-type` == 2)
edgelist_3 <- edgelist %>% filter(`tie-type` == 3)
edgelist_4 <- edgelist %>% filter(`tie-type` == 4)
edgelist_5 <- edgelist %>% filter(`tie-type` == 5)

# Tie type 1 plot
newnetwork_1 <- graph_from_data_frame(d = edgelist_1, vertices = nodelist, directed = TRUE)
plot(newnetwork_1, edge.arrow.size = 0.1, vertex.size = 3, 
     main = "Tie Type 1 Network - Messaged Last Week")

# Tie type 2 plot
newnetwork_2 <- graph_from_data_frame(d = edgelist_2, vertices = nodelist, directed = TRUE)
plot(newnetwork_2, edge.arrow.size = 0.1, vertex.size = 3, 
     main = "Tie Type 2 Network - Met During MBDS")

# Tie type 3 plot
newnetwork_3 <- graph_from_data_frame(d = edgelist_3, vertices = nodelist, directed = TRUE)
plot(newnetwork_3, edge.arrow.size = 0.1, vertex.size = 3, 
     main = "Tie Type 3 Network - Is My Friend")

# Tie type 4 plot
newnetwork_4 <- graph_from_data_frame(d = edgelist_4, vertices = nodelist, directed = TRUE)
plot(newnetwork_4, edge.arrow.size = 0.1, vertex.size = 3, 
     main = "Tie Type 4 Network - I Ask for Help")

# Tie type 5 plot
newnetwork_5 <- graph_from_data_frame(d = edgelist_5, vertices = nodelist, directed = TRUE)
plot(newnetwork_5, edge.arrow.size = 0.1, vertex.size = 3, 
     main = "Tie Type 5 Network - Asks Me for Help")
```

**Network Centrality Measures by Tie Type**

We then calculate measures centrality. We present results from centrality and
edge density, since their results are easier to interpret. 

Among the networks based on 5 tie types, type 4 (I ask for help) has 
the highest measures of centrality. Type 5 (Asks me for help) has
one of the lowest measures of centrality. 

```{r, include = F}
nodelist2 <- nodelist
# All centrality measures for overall network
degree(newnetwork, mode="in")
nodelist2['degree_closeness'] <- closeness(newnetwork, mode="all", weights=NA)
nodelist2['degree_betweeness'] <- betweenness(newnetwork, directed=T, weights=NA)

overall_centrality <- centr_degree(newnetwork, mode="in", normalized=T)
overall_density <- edge_density(newnetwork, loops=F)

# Centrality measures for tie1
degree(newnetwork_1, mode="in")
closeness(newnetwork_1, mode="all", weights=NA)
betweenness(newnetwork_1, directed=T, weights=NA)
centrality1 <- centr_degree(newnetwork_1, mode="in", normalized=T)
density1 <- edge_density(newnetwork_1, loops=F)

# Centrality measures for tie2
degree(newnetwork_2, mode="in")
closeness(newnetwork_2, mode="all", weights=NA)
betweenness(newnetwork_2, directed=T, weights=NA)
centrality2 <- centr_degree(newnetwork_2, mode="in", normalized=T)
density2 <- edge_density(newnetwork_2, loops=F)

# Centrality measures for tie3
degree(newnetwork_3, mode="in")
closeness(newnetwork_3, mode="all", weights=NA)
betweenness(newnetwork_3, directed=T, weights=NA)
centrality3 <- centr_degree(newnetwork_3, mode="in", normalized=T)
density3 <- edge_density(newnetwork_3, loops=F)

# Centrality measures for tie4
degree(newnetwork_4, mode="in")
closeness(newnetwork_4, mode="all", weights=NA)
betweenness(newnetwork_4, directed=T, weights=NA)
centrality4 <- centr_degree(newnetwork_4, mode="in", normalized=T)
density4 <- edge_density(newnetwork_4, loops=F)

# Centrality measures for tie5
degree(newnetwork_5, mode="in")
closeness(newnetwork_5, mode="all", weights=NA)
betweenness(newnetwork_5, directed=T, weights=NA)
centrality5 <- centr_degree(newnetwork_5, mode="in", normalized=T)
density5 <- edge_density(newnetwork_5, loops=F)
```


```{r, echo = F}
cat("Overall network centralization:", overall_centrality$centralization, "\n")
cat("Overall network edge density:", overall_density, "\n")

cat("Tie type 1 network centralization:", centrality1$centralization, "\n")
cat("Tie type 1 network edge density:", density1, "\n")

cat("Tie type 2 network centralization:", centrality2$centralization, "\n")
cat("Tie type 2 network edge density:", density2, "\n")

cat("Tie type 3 network centralization:", centrality3$centralization, "\n")
cat("Tie type 3 network edge density:", density3, "\n")

cat("Tie type 4 network centralization:", centrality4$centralization, "\n")
cat("Tie type 4 network edge density:", density4, "\n")

cat("Tie type 5 network centralization:", centrality5$centralization, "\n")
cat("Tie type 5 network edge density:", density5, "\n")
```


**Network Similarity Measures by Tie Type**

Next, we calculate assortativity to determine which network generates the most
similar network, most similar network based on introversion, and most
dissimilar network based on introversion.

Overall, type 1 network (texted last week) has high measure of similarity. 

Type 1 network also generates the most similar network in terms of introversion,
since it has the highest introversion-based assortativity value 0.1348994.

Tie Type 3 (is my friend) is the most dissimilar network in terms of introversion
since it has the lowest introversion-based assortativity value -0.02886641.
This means people are friends with each other regardless of whether they are
introverted or not.


```{r, echo = F}
# Similarity measures of overall network
node <- nodelist
node[is.na(node)] <- -1
net <- graph_from_data_frame(d = edgelist, vertices = node, directed = TRUE)
cat("Overall network assortativity:", assortativity_degree(newnetwork, directed=T) , "\n")
cat("Overall network assortativity based on Extraversion/Introversion:", 
    assortativity(newnetwork, V(net)$Extraversion, directed=T) , "\n")

cat("Tie type 1 network assortativity:", assortativity_degree(newnetwork_1, directed=T)  , "\n")
cat("Tie type 1 network assortativity based on Extraversion/Introversion:", 
    assortativity(newnetwork_1, V(net)$Extraversion, directed=T) , "\n")

cat("Tie type 2 network assortativity:", assortativity_degree(newnetwork_2, directed=T)  , "\n")
cat("Tie type 2 network assortativity based on Extraversion/Introversion:", 
    assortativity(newnetwork_2, V(net)$Extraversion, directed=T) , "\n")

cat("Tie type 3 network assortativity:", assortativity_degree(newnetwork_3, directed=T)  , "\n")
cat("Tie type 3 network assortativity based on Extraversion/Introversion:", 
    assortativity(newnetwork_3, V(net)$Extraversion, directed=T) , "\n")

cat("Tie type 4 network assortativity:", assortativity_degree(newnetwork_4, directed=T)  , "\n")
cat("Tie type 4 network assortativity based on Extraversion/Introversion:", 
    assortativity(newnetwork_4, V(net)$Extraversion, directed=T) , "\n")

cat("Tie type 5 network assortativity:", assortativity_degree(newnetwork_5, directed=T)  , "\n")
cat("Tie type 5 network assortativity based on Extraversion/Introversion:", 
    assortativity(newnetwork_5, V(net)$Extraversion, directed=T) , "\n")
```


# Question 3
*Do introverts tend to be at the periphery while extroverts are in the center of the network?*

Note: 1 means introverted, 2 means middle ground, 3 means extroverted, and -1 means
omitted response.

First, we calculated the average of closeness and betweenness grouped by level 
of extroversion. Higher betweeness means more connected and higher closeness 
means nodes are more closely connected. Based on our very limited sample,
introverted people in our class have high betweenness and low closeness, while 
extroverted people have low betweeness and slightly higher closeness.


We also graphed the overall network, colored by level of extroversion.
Based on the nodes of this graph, we can clearly see that, the very few people
who identify themselves as extroverted (3) are in the center of the network. 
Middle-ground people (2) are typically just outside the very center of the network 
Out of the three introverted individuals (1), two are also quite in the center 
of the network, and only one of them is rather on the 
periphery. 

Overall, it is  inconclusive if introverts tend to be on the periphery, while
extroverts are more likely to be in the center.
That said, our data set has very limited information, with
half of the respondents omitting answers to this question and only a few people
who identify themselves as either 1 or 3; the vast majority of people identified
themselves as middle ground. 


```{r, echo = F}
# Approach 1
extraversion_centrality <- nodelist2 %>% group_by(Extraversion) %>% 
  summarize(betweeness_mean = mean(degree_betweeness), closeness_mean = mean(degree_closeness))
print(extraversion_centrality) 
```

```{r, echo = F}
# Approach 2
plot(newnetwork,edge.arrow.size = 0.1,vertex.label=V(net)$Extraversion, 
     vertex.color = nodelist2$Extraversion)
#plot(subgraph(newnetwork, v = nodelist$ID), edge.arrow.size = 0.1, vertex.label=V(net)$Extraversion, vertex.color = nodelist2$Extraversion)
```




